Biogeography of crop progenitors and wild plant resources during the transition to agriculture in West Asia, 21--8.3 ka

### 2025-01-03 analysis/paper.qmd
Add ENM background

* Has always been important in study of agricultural origins
  * Historically: Vavilov, Pumpelly & Childe
  * Genetic studies tell us origin points, but not ranges
* Epipal./Neo. plant-based economies were diverse
  * More than the "founder crops"; 
  * More than food
  * (In archaebot., not all intentionally collected)
  * Regionally and temporally diverse
  * ...so we model lots of species!
* Regional ecological reconstructions generally rely on the 'expert interpolation' (or what do they call it with isoscapes?) method
  * See CSEAS (AEA-prep) presentation
  * Figure: comparisons

@@

* Review: @FranklinEtAl2015
* In SW Asia: @ConollyEtAl2012 (first use ever??), @CollinsEtAl2018
* Varying uses: to reconstruct human niches (esp. in Palaeolithic), to reconstruct spread of (esp. domestic) species from archaeological record, or to reconstruct wild ranges from modern occurrences and palaeoclimate (our approach)
* Caveats
  * In ENM literature, many sceptical about fore- and hind-casting
  * Archaeo data a golden opportunity here

>>>>

* Has always been important in study of agricultural origins
  * Historically: Vavilov, Pumpelly & Childe
  * Genetic studies tell us origin points, but not ranges
* Important to e.g.
  * Distinguish environmental from potentially anthropogenic change [@MartinEtAl2017; @MartinEtAl2025]
  * Reconstruct sequences of domestication [@YeomansEtAl2017]
* Epipal./Neo. plant-based economies were diverse
  * More than the "founder crops"; 
  * More than food
  * (In archaebot., not all intentionally collected)
  * Regionally and temporally diverse
  * ...so we model lots of species!
* Regional ecological reconstructions generally rely on the 'expert interpolation' (or what do they call it with isoscapes?) method
  * See CSEAS (AEA-prep) presentation
  * Figure: comparisons

@@

<!-- TODO: more citations? -->
Ecological niche modelling (ENM) or species distribution modelling (SDM) is widely used by ecologists to predict the geographic range of a species based on a set of environmental predictors.
Essentially, it involves combining records of where an organism has been observed with environmental data (climate, topography, etc.) for those locations to model the range of environmental values at which that species  – its environmental niche.
This model can then be used to predict the range of the organism in question either in the same or a different environment.
@CITE suggests reserving the term 'species distribution modelling' for when the method is used to recover the verifiable range of a species in a real and existing environment, and using 'ecological niche modelling' as the broader term covering hypothetical or predictive applications – a convention we follow here when referring to predictive or 'hindcast' models of past ranges.
Within this overarching framework, ecological niche modelling encompasses a wide range of applications and a variety of potential environmental predictors, modelling approaches, and methodologies, which we will not attempt to review here.

Ecological niche modelling has long been of interest to archaeologists as both a means of exploring the biological niche of humans and for reconstructing the past environments they inhabited [@DavidPollyEronen2011; @FranklinEtAl2015].
In the first sense, it has been used most extensively to model the range of humans and other hominin species [e.g. @BenitoEtAl2017; @YousefiEtAl2020; @BanksEtAl2021; @YaworskyEtAl2024a; @YaworskyEtAl2024b; @GuranEtAl2024], especially in the Palaeolithic.
This overlaps with what archaeologists usually call generically 'predictive modelling' [@VerhagenWhitley2020]—more precisely 'site distribution modelling'—which is essentially the same approach as (and often borrows methodologies from) ecological niche modelling but applied to the occurrence of archaeological sites.
Here what is modelled is not strictly a biological niche alone, but also aspects of human geography, taphonomy, and archaeological visibility.
These applications can be distinguihed from 'palaeoecological niche modelling', where the object of model remains, as in ecology, a non-human biological niche.

@FranklinEtAl2015 review palaeoecological niche modelling and advocate for its greater adoption in environmental archaeology.
One relevant early example is @ConollyEtAl2012 used the occurrence of wild and domestic *Bos* remains at prehistoric archaeological sites in Europe and West Asia to map the evolving niche of cattle over the Pleistocene–Holocene transition.
It has been used to model the availability of fauna exploited by humans at wider scales [e.g. @deAndresHerreroEtAl2018; @YaworksyEtAl2023] and, in a West Asian context, of foraged plant resources in the landscape around the Neolithic site of XX<!-- TODO: Kortik? --> [@CollinsEtAl2018].
Modelling the spread of crops has been another significant archaeological application [@CremaEtAl].

In the majority of studies to date (palaeo)ecological niche modelling has been applied to archaeological data in an 'inductive' fashion, i.e. faunal and botanical remains from ancient sites are used as the occurrence dataset for training a model using either past or present environmental data.
However, both the zooarchaeological and archaeobotanical records are sparse and subject to a complex array of depositional, taphonomic and recovery biases factors that , many of which are not fully understood and/or cannot be corrected for.
This means that while the archaeological attestation of the presence of a species might generally be relied upon, it is highly unlikely that its absence is representative of true past distributions.

The alternative approach is to train the model using contemporary occurrence and environmental data and then use palaeoenvironmental data to 'hindcast' its predictions backwards in time.
Like @FranklinEtAl2015, we view the hindcasting approach as more promising, because training datasets for both occurrences and environment are far more readily available, complete and reliable for the present than the past.
There is some scepticism in the ecological niche modelling literature about the ability of such models to make accurate predictions in unknown environments (like the past) [@CITES?], 
but here the hindcasting approach also presents an opportunity:
it reserves archaeological occurrence data as an independent dataset that can be used to assess the retrodictive performance of the model.
This possibily was suggested by @FranklinEtAl2015 but to our knowledge our study represents the first attempt to actually do so.

The major practical limitation of the hindcasting approach is that it relies on spatially explicit, high resolution palaeoenvironmental surfaces with continuous coverage of the region and periods of interest.
Until recently, this has not been widely available for most applications, which is perhaps why only a minority of studies use it [cf. @YaworskyEtAl2023].
In this study, we are able to take advantage of the increasing availability of high resolution, global palaeoclimate data derived from simulation experiments with general circulation models of climate [@BrownEtAl2018; @BrownEtAl2020; @KargerEtAl2023].
<!-- TODO: mention that GBIF, CHELSA, SoilGrids, WorldClim, DEMs, etc. are also great? -->

### 2025-01-03 analysis/paper.tex
Add ENM background

  \begin{itemize}
  \tightlist
  \item
    Historically: Vavilov, Pumpelly \& Childe
  \item
    Genetic studies tell us origin points, but not ranges
  \end{itemize}
\item
  Epipal./Neo. plant-based economies were diverse

@@

\begin{itemize}
\tightlist
\item
  Review: (Franklin et al., 2015)
\item
  In SW Asia: (Conolly et al., 2012) (first use ever??), (Collins et
  al., 2018)
\item
  Varying uses: to reconstruct human niches (esp.~in Palaeolithic), to
  reconstruct spread of (esp.~domestic) species from archaeological
  record, or to reconstruct wild ranges from modern occurrences and
  palaeoclimate (our approach)
\item
  Caveats

  \begin{itemize}
  \tightlist
  \item
    In ENM literature, many sceptical about fore- and hind-casting
  \item
    Archaeo data a golden opportunity here
  \end{itemize}
\end{itemize}

>>>>

  \begin{itemize}
  \tightlist
  \item
    Historically: Vavilov, Pumpelly \& Childe
  \item
    Genetic studies tell us origin points, but not ranges
  \end{itemize}
\item
  Important to e.g.

  \begin{itemize}
  \tightlist
  \item
    Distinguish environmental from potentially anthropogenic change
    (\textbf{MartinEtAl2017?}; \textbf{MartinEtAl2025?})
  \item
    Reconstruct sequences of domestication (\textbf{YeomansEtAl2017?})
  \end{itemize}
\item
  Epipal./Neo. plant-based economies were diverse

@@

Ecological niche modelling (ENM) or species distribution modelling (SDM)
is widely used by ecologists to predict the geographic range of a
species based on a set of environmental predictors. Essentially, it
involves combining records of where an organism has been observed with
environmental data (climate, topography, etc.) for those locations to
model the range of environmental values at which that species -- its
environmental niche. This model can then be used to predict the range of
the organism in question either in the same or a different environment.
(\textbf{CITE?}) suggests reserving the term `species distribution
modelling' for when the method is used to recover the verifiable range
of a species in a real and existing environment, and using `ecological
niche modelling' as the broader term covering hypothetical or predictive
applications -- a convention we follow here when referring to predictive
or `hindcast' models of past ranges. Within this overarching framework,
ecological niche modelling encompasses a wide range of applications and
a variety of potential environmental predictors, modelling approaches,
and methodologies, which we will not attempt to review here.

Ecological niche modelling has long been of interest to archaeologists
as both a means of exploring the biological niche of humans and for
reconstructing the past environments they inhabited (Franklin et al.,
2015; \textbf{DavidPollyEronen2011?}). In the first sense, it has been
used most extensively to model the range of humans and other hominin
species (e.g. \textbf{BenitoEtAl2017?}; \textbf{YousefiEtAl2020?};
\textbf{BanksEtAl2021?}; \textbf{YaworskyEtAl2024a?};
\textbf{YaworskyEtAl2024b?}; \textbf{GuranEtAl2024?}), especially in the
Palaeolithic. This overlaps with what archaeologists usually call
generically `predictive modelling'
(\textbf{VerhagenWhitley2020?})---more precisely `site distribution
modelling'---which is essentially the same approach as (and often
borrows methodologies from) ecological niche modelling but applied to
the occurrence of archaeological sites. Here what is modelled is not
strictly a biological niche alone, but also aspects of human geography,
taphonomy, and archaeological visibility. These applications can be
distinguihed from `palaeoecological niche modelling', where the object
of model remains, as in ecology, a non-human biological niche.

(Franklin et al., 2015) review palaeoecological niche modelling and
advocate for its greater adoption in environmental archaeology. One
relevant early example is (Conolly et al., 2012) used the occurrence of
wild and domestic \emph{Bos} remains at prehistoric archaeological sites
in Europe and West Asia to map the evolving niche of cattle over the
Pleistocene--Holocene transition. It has been used to model the
availability of fauna exploited by humans at wider scales (e.g.
\textbf{deAndresHerreroEtAl2018?}; \textbf{YaworksyEtAl2023?}) and, in a
West Asian context, of foraged plant resources in the landscape around
the Neolithic site of XX (Collins et al., 2018). Modelling the spread of
crops has been another significant archaeological application
(\textbf{CremaEtAl?}).

In the majority of studies to date (palaeo)ecological niche modelling
has been applied to archaeological data in an `inductive' fashion,
i.e.~faunal and botanical remains from ancient sites are used as the
occurrence dataset for training a model using either past or present
environmental data. However, both the zooarchaeological and
archaeobotanical records are sparse and subject to a complex array of
depositional, taphonomic and recovery biases factors that , many of
which are not fully understood and/or cannot be corrected for. This
means that while the archaeological attestation of the presence of a
species might generally be relied upon, it is highly unlikely that its
absence is representative of true past distributions.

The alternative approach is to train the model using contemporary
occurrence and environmental data and then use palaeoenvironmental data
to `hindcast' its predictions backwards in time. Like (Franklin et al.,
2015), we view the hindcasting approach as more promising, because
training datasets for both occurrences and environment are far more
readily available, complete and reliable for the present than the past.
There is some scepticism in the ecological niche modelling literature
about the ability of such models to make accurate predictions in unknown
environments (like the past) (\textbf{CITES?}), but here the hindcasting
approach also presents an opportunity: it reserves archaeological
occurrence data as an independent dataset that can be used to assess the
retrodictive performance of the model. This possibily was suggested by
(Franklin et al., 2015) but to our knowledge our study represents the
first attempt to actually do so.

The major practical limitation of the hindcasting approach is that it
relies on spatially explicit, high resolution palaeoenvironmental
surfaces with continuous coverage of the region and periods of interest.
Until recently, this has not been widely available for most
applications, which is perhaps why only a minority of studies use it
(cf. Yaworsky et al., 2023). In this study, we are able to take
advantage of the increasing availability of high resolution, global
palaeoclimate data derived from simulation experiments with general
circulation models of climate (Brown et al., 2018;
\textbf{BrownEtAl2020?}; \textbf{KargerEtAl2023?}).
